A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-06 latex/main.tex
Updates from Overleaf

The most comprehensive analysis on recombination in SARS-CoV-2 to 
date has been performed using the combination of UShER and 
RIPPLES~\cite{Turakhia2022}. 
% NOTE: not citing RIVET here as there is no real analysis
% in the paper itself, it's just focused on the tool
RIPPLES works by taking an UShER phylogeny and systematically breaking up
sequences subtended by long branches into multiple segments 
that are placed independently.
A recombination event at a node is flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny.
This approach has much in common with the LS HMM
used in sc2ts, but with one important distinction.
While sc2ts detects and incorporates recombination
in ``real-time'' as sequences are added to the ARG in daily
batches, RIPPLES operates on a complete phylogeny,
attempting to detect recombination post-hoc among millions of nodes. 
Like any phylogenetic reconstruction method
that ignores recombination, UShER must force placement of recombinant
sequences in places that minimize the discrepancy across the entire sequence,
which may not be near any of the true parents. 
These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
may break up the artefactual long branches used to detect recombination.

We compared the performance of sc2ts recombinant detection 
against RIPPLES run on the Viridian UShER phylogeny,
subset to the 2,475,418 samples shared with the ARG (STAR Methods).
To enable exact comparison, we omitted the QC filtering steps from both
methods, and initially set the RIPPLES ``parsimony improvement'' 
parameter to $p=4$ (rather than the more lenient default of 3), 
to match the equivalent $k=4$ setting used in sc2ts (STAR Methods). 
Under this setting,
RIPPLES reported 1174 recombination events of which 
601 were validated by 3SEQ (51\%; STAR Methods).
% JK: There's no point in making a meal of the 3SEQ validation rate,
% This is notably lower than the equivalent 
% concordance rate of 57\% among the 855 recombination events reported by sc2ts.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both,
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination of the choice of sites 
used for inference, and the strict time ordering imposed
by sc2ts (Document~\ref{sec:detailed_ripples}).
The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterised events 
associated with the Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows the events detected 
by UShER+RIPPLES associated with samples assigned to Pango X
lineages, showing that at $p=4$
only 9 of the Pango X lineages are present and
with unequivocal recombinant origin for 6 
(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%;
Figure~\ref{fig:ripples_sc2ts_events}).
At this more lenient threshold, 23 of the 44 Pango X lineages
are present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages associated with two competing events 
(Document~\ref{sec:detailed_ripples}).
XB and XAJ are among the lineages detected by UShER+RIPPLES,
strengthening evidence that these are false negatives 
for sc2ts (Table~\ref{tab:pango_x_lineages}).

The choice of the $p$ threshold has a large effect 
on the number of recombination events detected by RIPPLES, and 
previous studies have used the more lenient (default) value of 
$3$\cite{Turakhia2022,Smith2023tracking}.
For example, when applied to an UShER phylogeny of
$\sim$1.6 million genomes sampled up to May, 2021,
RIPPLES reported 589 unique recombination events with 43,104 descendant
samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
Over the same time interval, the sc2ts ARG ($k=4$) identifies
14 recombination events passing QC with
172 descendant samples among 501,235 samples (0.034\%),
a 40- to 80-fold reduction.
The sc2ts ARG identifies more of the well-characterised Pango X lineages
at a more stringent threshold, and therefore provides us with a 
conservative lower bound
on the frequency of detectable recombination events in SARS-CoV-2.

@@

\subsection{Detailed RIPPLES analysis}
\label{sec:detailed_ripples}
We ran RIPPLES and aligned events with the sc2ts ARG as described in STAR
Methods. We begin by describing the overall composition of events output 
by RIPPLES at the two parameter values used.

At $p=4$, RIPPLES returned a total of 1174 events, and 
610 pass 3SEQ validation (51\%). 1052 events
(89.6\%) were in close clade agreement with the sc2ts ARG,
and 468 (40\%) corresponded to recombination events in the ARG
(Figure~\ref{fig:ripples_sc2ts_events}A).
We investigated the 701 RIPPLES events that could be aligned 
with the ARG but were not close to a recombination node
to determine why they were not found by sc2ts.
For 408 of these RIPPLES identified recombination events,
one or more parents were \emph{younger} than the
recombinant sequence based on metadata,
and therefore impossible under the strict time ordering 
enforced by sc2ts.
Secondly, of the 179
remaining events with matching clades and in which the 
sc2ts node has more than 4 mutations, of the total
1,167 mutations, 452 of these occurred at sites that were 
remapped during post-hoc parsimony (and thus the mutations
were not observed by the LS HMM during primary inference). 
The remaining ``missing'' events are likely attributable 
to the finer details of the 
differences in tree building approaches between UShER and sc2ts.

At $p=3$, RIPPLES returned a total of 4111 events, and 
2718 pass 3SEQ validation (66\%). 3755 events
(91.3\%) were in close clade agreement with the sc2ts ARG,
and 595 (14\%) corresponded to recombination events in the ARG
(Figure~\ref{fig:ripples_sc2ts_events}B).
We did not investigate the $p=3$ events any further as the additional
events detected at the lower threshold are not expected to be detected
by sc2ts at the parameter value used ($k=4$).

\begin{table}
\caption{
Distribution of numbers of descendants for UShER+RIPPLES events
and sc2ts. The first column shows the number of descendants (binned
for larger values) and the others show the corresponding number of events.
}
\label{tab:ripples_sc2ts_event_descendants}
\centering
\begin{tabular}{lrrr}
\toprule
Count & RIPPLES p3 & RIPPLES p4 & sc2ts \\
\midrule
1 & 2616 & 757 & 525 \\
2 & 575 & 178 & 124 \\
3 & 232 & 65 & 47 \\
4 & 124 & 38 & 32 \\
5 & 74 & 14 & 17 \\
5-100 & 437 & 108 & 96 \\
100-1000 & 32 & 6 & 8 \\
1000-10k & 15 & 6 & 4 \\
10k-100k & 2 & 0 & 0 \\
100k-1m & 4 & 2 & 2 \\
\bottomrule
\end{tabular}
\end{table}

Next we considered the overall distribution of the numbers of descendants
of the events returned by sc2ts and RIPPLES 
(Table~\ref{tab:ripples_sc2ts_event_descendants}).
Overall, the distributions were similar, with the majority having 
very few descendants and being dominated by singletons.
The events subtending $>$ 100,000 samples
in all three cases are noteworthy (see
Document~\ref{sec:spurious-ancestrally-important-recombinants} 
for analysis of the two sc2ts events).

@@

\begin{tabular}{lrrlrl}
\toprule
sc2ts event & in $p=4$ & UShER desc & sc2ts class & clade diff & descendants \\
\midrule
\multicolumn{6}{l}{Equivalent Sc2ts and RIPPLES recombination events}\\
XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
XBR & $\checkmark$ & 1 & I & 0 & {XBR:1} \\
% XM3 & $\checkmark$ & 1 & I & 0 & {XM:1} \\
XS & $\checkmark$ & 17 & I & 0 & {XS:17} \\
XA & $\checkmark$ & 39 & I & 0 & {XA:39} \\
XY &  & 23 & I & 0 & {XY:23} \\
XW &  & 32 & I & 0 & {XW:32} \\
XL &  & 64 & I & 0 & {XL:64} \\
XBB &  & 6452 & I & 1 & {XBB:6452} \\
\midrule
\multicolumn{6}{l}{RIPPLEs recombination event close to sc2ts recombination event}\\
XBM & $\checkmark$ & 12 & II & 0 & {XBM:10, BF.3:2} \\
XM & $\checkmark$ & 8 & II & 0 & {XM:4, BA.2:4} \\
Xx & $\checkmark$ & 18 & II & 0 & {XAC:18} \\
Xx &  & 9 & II & 0 & {XAE:9} \\
XQ &  & 1 & II & 0 & {XAM:1} \\
XQ &  & 21 & II & 0 & {XAM:21} \\
XE/XH &  & 1113 & II & 3 & {XE:1113} \\
\midrule
\multicolumn{6}{l}{RIPPLEs recombination event equivalant to sc2ts non-recombination event}\\
XB & $\checkmark$ & 411 & IV & 0 & {XB:192, B.1:219} \\
\midrule
\multicolumn{6}{l}{RIPPLEs recombination event not comparable to sc2ts}\\
NA & $\checkmark$ & 19 & NA & 45190 & {XAJ:18, BA.4:1} \\
NA & $\checkmark$ & 35 & NA & 7147 & {XBD:30, BA.2.75:5} \\
NA & $\checkmark$ & 36 & NA & 7146 & {XBR:1, XBD:30, BA.2.75:5} \\
NA &  & 20 & NA & 1061775 & {XS:17, BA.1:1, BA.1.1:1, BA.1.15:1} \\
NA &  & 1321 & NA & 340501 & {XE:1113, XJ:68, BA.2:82, XM:29, XY:23,
XAL:3, XAF:1, XH:2} \\
\bottomrule
\end{tabular}

@@

We next examined the RIPPLES events referencing Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows a summary of all
recombination events associated with Pango X lineages detected by RIPPLES
at $p=3$ and their relationship with the sc2ts ARG.
Of the 16 Class I events (X lineages unequivocally associated with
a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
8 have corresponding recombination events (4, at $p=4$).
Of the 5 Type II events
(recombination nodes with more complex assocation with X lineages),
two have close matches.
XBM matches exactly (including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH samples in the clade).
UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
makes these both descendants of the Xx event,
RIPPLES creates two independent recombinations.
Similarly, XAC and XAE have exactly matching clades, but RIPPLES
infers two independent recombination events while sc2ts groups them both
under the Xx event. XB is notable as a likely sc2ts false negative that is 
identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
associated with X lineages output by RIPPLES cannot be easily mapped 
to the sc2ts ARG as one or more descendants is distantly placed.
XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 
non-recombinant with the largest number of mutations 
(Table~\ref{tab:pango_x_lineages}), and a likely false negative.
The remaining 4 events are mixtures of multiple lineages that are 
difficult to reconcile with either the other events in the table 
or the sc2ts ARG.

See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/analysis\_ripples.ipynb}
{analysis\_ripples.ipynb}
for details.

\subsection{Likely spurious recombinants with many descendants}
\label{sec:spurious-ancestrally-important-recombinants}
In the sc2ts ARG, there are two recombination events which have a large number of
descendants. The first of these
corresponds to recombination event within BA.2 whose breakpoint is at position 27382
on the far right of the genome.
% JK not sure what this is referring to?
% (see supplementary document S3, node 1030562). 
Although
supported by 5 sites, three of them are adjacent so that they only represent 3 loci,
meaning that this event does not pass our QC filter..

The second of these involves a single breakpoint at position 26858, and is the origin
of the BA.5 lineage in the ARG.
% (see supplementary document S3, node 1189192). 
Although supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
some of which are adjacent to the supporting loci. Furthermore, when the first BA.5
sample is added to the ARG, it is possible to construct a more parsimonious
non-recombinant origin by long branch splitting, 
although subsequent tree building does not make this a parsimonious 
way to rewire the final ARG.  It is therefore likely to be spurious.

>>>>

The most comprehensive analysis on recombination in SARS-CoV-2 to 
date has been performed using the combination of UShER and 
RIPPLES~\cite{Turakhia2022}. 
% NOTE: not citing RIPPLES here as there is no real analysis
% in the paper itself, it's just focused on the tool
RIPPLES works by taking a
pre-existing UShER phylogeny and systematically breaking up
sequences subtended by long branches into multiple segments 
that are placed independently.
A recombination event at a node is flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny.
This approach has much in common with the LS HMM
used in sc2ts, but with one important distinction.
While sc2ts detects and incorporates recombination
in ``real-time'' as sequences are added to the ARG in daily
batches, RIPPLES operates on a complete phylogeny,
attempting to detect recombination post-hoc among millions of nodes. 
Like any phylogenetic reconstruction method
that ignores recombination, UShER must force placement of recombinant
sequences in places that minimize the discrepancy across the entire sequence,
which may not be near any of the true parents. 
These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
may break up the artefactual long branches used to detect recombination.

We compared the performance of sc2ts recombinant detection against RIPPLES run on the Viridian UShER phylogeny,
subset to the 2,475,418 samples shared with the ARG (STAR Methods).
To enable exact comparison, we omitted the QC filtering steps from both
methods, and initially set the RIPPLES ``parsimony improvement'' parameter to $p=4$ (rather than the more lenient default of 3), to match the equivalent $k=4$ setting
used in sc2ts (STAR Methods). Under this setting,
RIPPLES reported 1174 recombination events of which 601 were designated as recombinant when running 3SEQ on the candidate trios (STAR Methods), giving a concordance rate of 51\%. This is notably lower than the equivalent concordance rate of 57\% among the 855 recombination events reported by sc2ts.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both,
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination of the choice of sites 
used for inference, and the strict time ordering imposed
by sc2ts (Document~\ref{sec:detailed_ripples}).
The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterised events 
associated with the Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows the events detected 
by UShER+RIPPLES associated with samples assigned to Pango X
lineages, showing that at $p=4$
only 9 of the Pango X lineages are present and
with unequivocal recombinant origin for 6 
(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%;
Figure~\ref{fig:ripples_sc2ts_events}).
At this more lenient threshold, 23 of the 44 Pango X lineages
are present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages associated with two competing events 
(Document~\ref{sec:detailed_ripples}).
XB and XAJ are among the lineages detected by UShER+RIPPLES,
strengthening evidence that these are false negatives 
for sc2ts (Table~\ref{tab:pango_x_lineages}).

The choice of the $p$ threshold has a large effect 
on the number of recombination events detected by RIPPLES, and 
previous studies have used the more lenient (default) value of 
$3$\cite{Turakhia2022,Smith2023tracking}.
For example, when applied to an UShER phylogeny of
$\sim$1.6 million genomes sampled up to May, 2021,
RIPPLES reported 589 unique recombination events with 43,104 descendant
samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
Over the same time interval, the sc2ts ARG ($k=4$) identifies
14 high-confidence recombination events with
172 descendant samples among 501,235 samples (0.034\%),
a 40- to 80-fold reduction.
The sc2ts ARG identifies more of the well-characterised Pango X lineages
at a more stringent threshold, and therefore provides us with a 
conservative lower bound
on the frequency of detectable recombination events in SARS-CoV-2.

@@

\subsection{Detailed RIPPLES analysis}
\label{sec:detailed_ripples}

Using the default ``parsimony improvement'' parameter of $p=3$, UShER+RIPPLES detected a total of 4111 recombination events.
Most of these subtend $<10$ samples (2616 singletons),
although four events subtended more than 100,000 samples.
Analysis of the descending samples for each event shows that
10 of the 44 Pango X lineages present in the dataset
% NOTE: listing them like this for sorting and manipulation purposes
(%
XA, 
XB, 
XC, 
XL, 
XW,
XAC, 
XAE, 
XAJ,
XBB, 
XBM%
)
are associated unequivocally with a single recombination event.
7 lineages 
(%
XE,
XM,
XS,
XY,
XAM,
XBD, 
XBR%
)
are associated with two events. Finally, 
one event subtending 1321 samples suggests that 
XE, 
XH, 
XJ, 
XM, 
XY, 
XAF,
XAL, 
and 
82 BA.2 samples share a recombination event whose parents are BA.2 and BA.1 [FIXME: With what breakpoint and parents?].
We also ran RIPPLES with $p=4$, which reduced the overall number of
events to 1174 (roughly 4-fold) but also approximately halved the number of
events involving X lineages, missing well-characterized events,
such as XBB (Table~\ref{tab:ripples_x_lineages}).

To compare these results with the sc2ts ARG, we first matched clades
by finding the most recent common ancestor (MRCA) of the samples
descending from each RIPPLES event. Overall, clade agreement was
high with [FIXME, get exact numbers] 37XX nodes (90\%) in the sc2ts ARG
where descendants match exactly. Of these nodes, 68\% had no recombinant
ancestry. Of the remainder, 595 could be matched with
one of the 855 recombination events in the sc2ts ARG
(Figure~\ref{fig:ripples_sc2ts_events}).
Table~\ref{tab:ripples_x_lineages} shows a summary of
all recombination events associated with Pango X lineages detected by RIPPLES,
and their relationship with the sc2ts ARG.
Of the 16 Class I events (Pango X lineages unequivocally associated with
a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
8 are found by RIPPLES (4, at $p=4$).
Of the 5 Type II events
(recombination nodes with more complex assocation with X lineages),
two have close matches in RIPPLES, with XBM matching exactly
(including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH samples in the clade).
Two events in the RIPPLES set correspond to likely false negatives
in sc2ts: XB (Document~\ref{sec:dd_xb})
and XAJ (Document~\ref{sec:dd_xaj}).

Firstly, 408 of these RIPPLES identified recombination events
have one or more parents that are \emph{younger} than the
recombinant sequence based on metadata. As sc2ts enforces strict time ordering,
the parental sequences were simply not available for it to match against
when the recombinant sequence was inserted. Secondly, of the 179
remaining events with matching clades and in which the 
sc2ts node has more than 4 mutations, of the total
1,167 mutations, 452 of these occurred at sites that were 
remapped during post-hoc parsimony. The remaining ``missing''
events are likely attributable to the finer details of the 
differences in tree building approaches between UShER and sc2ts.

[
FIXME: Discuss the large effect recombinants inferred by both in more detail. E.g.:
sc2ts identifies two recombination events with \textgreater 10k descendants (specifically, 166,979 and 166,285 descendant samples),
both of which we identify as spurious (see \ref{sec:spurious-ancestrally-important-recombinants}).
RIPPLES at $p=4$ also identifies two recombination events with \textgreater 10k descendants (531,224 and 339,523 descendant samples);
these do not correspond to the same two spurious large-effect events in sc2ts. At $p=3$, RIPPLES identifies six recombination events with \textgreater 10k descendant samples, four of which have \textgreater 100k descendant samples.
]

@@

\begin{tabular}{lrrlrl}
\toprule
sc2ts event & in p4 & UShER desc & sc2ts class & clade diff & descendants \\
\midrule
\multicolumn{6}{l}{Equivalent Sc2ts and RIPPLEs recombination events}\\
XBR & $\checkmark$ & 1 & I & 0 & {XBR:1} \\
% XM3 & $\checkmark$ & 1 & I & 0 & {XM:1} \\
XC & $\checkmark$ & 5 & I & 0 & {XC:5} \\
XS & $\checkmark$ & 17 & I & 0 & {XS:17} \\
XY &  & 23 & I & 0 & {XY:23} \\
XW &  & 32 & I & 0 & {XW:32} \\
XA & $\checkmark$ & 39 & I & 0 & {XA:39} \\
XL &  & 64 & I & 0 & {XL:64} \\
XBB &  & 6452 & I & 1 & {XBB:2361} \\
XBM & $\checkmark$ & 12 & II & 0 & {XBM:10, BF.3:2} \\
\midrule
\multicolumn{6}{l}{RIPPLEs recombination event close to sc2ts recombination event}\\
XE/XH &  & 1113 & II & 3 & {XE:1113} \\
XQ &  & 1 & III & 0 & {XAM:1} \\
XQ &  & 21 & III & 0 & {XAM:21} \\
Xx &  & 9 & III & 0 & {XAE:9} \\
Xx & $\checkmark$ & 18 & III & 0 & {XAC:18} \\
XM & $\checkmark$ & 8 & III & 0 & {BA.2:4, XM:4} \\
\midrule
\multicolumn{6}{l}{RIPPLEs recombination event equivalant to sc2ts non-recombination event}\\
XB & $\checkmark$ & 411 & IV & 0 & {B.1:219, XB:192} \\
\midrule
\multicolumn{6}{l}{RIPPLEs recombination event not comparable to sc2ts}\\
NA & $\checkmark$ & 19 & NA & 45190 & {XAJ:18, BA.4:1} \\
NA &  & 20 & NA & 1061775 & {BA.1.15:1, XS:17, BA.1:1, BA.1.1:1} \\
NA & $\checkmark$ & 35 & NA & 7147 & {XBD:30, BA.2.75:5} \\
NA & $\checkmark$ & 36 & NA & 7146 & {XBD:30, XBR:1, BA.2.75:5} \\
NA &  & 1321 & NA & 340501 & {XE:1113, XH:2, XJ:68, XY:23, BA.2:82, XM:29,
XAL:3, XAF:1} \\
\bottomrule
\end{tabular}

@@

[FIXME: this is an earlier draft of main text which explains the table in
detail]

We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
We first ran RIPPLES with parameter $p=4$, as being roughly comparable
to the recombination penalty used in sc2ts ($k=4$; STAR Methods).
This resulted in 1,174 recombination events reported, of which 460 could be mapped
to one of the 855 recombination events in the sc2ts ARG (STAR Methods;
Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
samples showed that many Pango X lineages were not associated with
recombination events (including XBB), and we therefore ran
RIPPLES with the more lenient value of $p=3$ (the default).
This resulted in a total of 4111 recombination events
of which 595 have corresponding recombination events in the 
sc2ts ARG (Figure \ref{fig:ripples_sc2ts_events}B).

Table~\ref{tab:ripples_x_lineages} shows a summary of all
recombination events associated with Pango X lineages detected by RIPPLES,
and their relationship with the sc2ts ARG.
Of the 16 Class I events (Pango X lineages unequivocally associated with
a recombination node in the sc2ts ARG) in Table~\ref{tab:pango_x_lineages},
8 have corresponding recombination events (4, at $p=4$).
Of the 5 Type II events
(recombination nodes with more complex assocation with X lineages),
two have close matches.
XBM matches exactly (including 2 descending BF.3 samples),
and XE/XH is similar (but without including the 2 XH samples in the clade).
UShER and sc2ts agree on creating two separate XAM clades, but while sc2ts
makes these both descendants of the Xx event,
RIPPLES creates two independent recombinations.
Similarly, XAE and [FIXME] ??? have exactly matching clades, but RIPPLES
infers two independent recombination events while sc2ts groups them both
under the Xx event. XB is notable as a likely sc2ts false negative that is 
identified by RIPPLES at both $p=3$ and $p=4$. The remaining events 
associated with X lineages output by RIPPLES cannot be easily mapped 
to the sc2ts ARG as one or more descendants is distantly placed.
XAJ is an outlier for sc2ts, being the Pango X lineage identified as a 
non-recombinant with the largest number of mutations 
(Table~\ref{tab:pango_x_lineages}), and a likely false negative.
The remaining 4 events are mixtures of multiple lineages that are 
difficult to reconcile with either the other events in the table 
or the sc2ts ARG.

\subsection{Likely spurious recombinants with many descendants}
\label{sec:spurious-ancestrally-important-recombinants}
In the sc2ts ARG, there are two recombination events which have a large number of
descendants but which are not classified as recombinants by Pango. The first of these
corresponds to recombination event within BA.2 whose breakpoint is at position 27382
on the far right of the genome (see supplementary document S3, node 1030562). Although
supported by 5 sites, three of them are adjacent so that they only represent 3 loci,
meaning that this event does not pass our recombinant quality control measure.

The second of these involves a single breakpoint at position 26858, and is the origin
of the BA.5 lineage in the ARG (see supplementary document S3, node 1189192). Although
supported by 5 loci on the right, this recombinant also requires 3 de-novo mutations,
some of which are adjacent to the supporting loci. Furthermore, when the first BA.5
sample is added to the ARG, it is possible to construct a more parsimonious
non-recombinant origin by long branch splitting, 
although subsequent tree building does not make this a parsimonious
way to rewire the final ARG.
